---
title: "NAFLD Transcriptomic Analysis"
author: "Nicolas Legrand"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---


## 1. Dataset loading and initial exploration
# Load GEO dataset and inspect expression matrix + metadata

```{r setup}
# ---- Setup ----
library(GEOquery)
library(tidyverse)

# ---- Download GEO dataset ----
gse <- getGEO("GSE135251", GSEMatrix = TRUE)

# Verify download
length(gse)
names(gse)

expr <- exprs(gse[[1]])
meta <- pData(gse[[1]])

dim(expr)
head(expr[,1:5])
head(meta)

```

## 2. Download and extract raw count files
# Retrieve supplementary files and extract raw featureCounts outputs

```{r download raw}
library(GEOquery)

# Télécharger les fichiers supplémentaires (RAW)
getGEOSuppFiles("GSE135251")

# Décompresser l’archive RAW dans un sous-dossier
untar("GSE135251/GSE135251_RAW.tar", exdir = "GSE135251/raw")

# Lister les fichiers disponibles
list.files("GSE135251/raw")

```
## 3. Merge raw count files into a single count matrix
# Read all featureCounts files and merge them by Geneid

```{r merge_counts}
# ---- merge_counts_fixed ----

library(tidyverse)

files <- list.files("GSE135251/raw", full.names = TRUE)

read_count_file <- function(f) {
  df <- read.delim(f, header = TRUE)
  
  # Identify gene column (Geneid or first column)
  gene_col <- grep("Geneid|gene|Gene|ID", colnames(df), ignore.case = TRUE)
  if (length(gene_col) == 0) gene_col <- 1
  
  # Identify count column (usually second column)
  count_col <- grep("count|Counts|read", colnames(df), ignore.case = TRUE)
  if (length(count_col) == 0) count_col <- 2
  
  df <- df[, c(gene_col[1], count_col[1])]
  colnames(df) <- c("Geneid", basename(f))
  
  return(df)
}

# Read all files
count_list <- lapply(files, read_count_file)

# Merge all by Geneid
counts <- Reduce(function(x, y) merge(x, y, by = "Geneid", all = TRUE), count_list)

# Replace NA by 0
counts[is.na(counts)] <- 0

# Set rownames
rownames(counts) <- counts$Geneid
counts <- counts[, -1]

# Check
dim(counts)
counts[1:5, 1:5]

```
## 4. Clean count matrix
# Remove featureCounts summary rows and convert to integer matrix

```{r clean_counts}
# ---- clean_counts ----

# Remove featureCounts summary rows
bad_rows <- grep("^__", rownames(counts))
length(bad_rows)   # should be 5

counts_clean <- counts[-bad_rows, ]

# Convert to integer matrix (DESeq2 requirement)
counts_clean <- as.matrix(counts_clean)
mode(counts_clean) <- "integer"

# Check again
dim(counts_clean)
counts_clean[1:5, 1:5]

```

## 5. Clean and format metadata
# Select relevant columns and create the condition factor

```{r clean_metadata}

# ---- clean_metadata ----

library(dplyr)
library(stringr)

# Start from meta
meta_clean <- meta %>%
  dplyr::select(`group in paper:ch1`, `fibrosis stage:ch1`, `nas score:ch1`)

# Rename columns for clarity
meta_clean <- meta_clean %>%
  dplyr::rename(
    group = `group in paper:ch1`,
    fibrosis_stage = `fibrosis stage:ch1`,
    nas_score = `nas score:ch1`
  )

# Extract condition from group
meta_clean$condition <- meta_clean$group

# Make condition a factor
meta_clean$condition <- factor(meta_clean$condition)

# Check
table(meta_clean$condition)
head(meta_clean)


```

```{r clean_colnames}
# ---- clean_colnames ----

# Extract GSM ID from filenames
clean_names <- colnames(counts_clean) %>%
  stringr::str_extract("GSM[0-9]+")

# Replace column names
colnames(counts_clean) <- clean_names

# Check
head(colnames(counts_clean))

```
## 6. Align metadata with count matrix
# Ensure sample names match between counts and metadata

```{r check_alignment}
# ---- check_alignment2 ----

# Check again
all(colnames(counts_clean) == rownames(meta_clean))

# Align metadata to counts
meta_aligned <- meta_clean[colnames(counts_clean), ]

# Check
all(colnames(counts_clean) == rownames(meta_aligned))

head(meta_aligned)

```

## 7. Create DESeq2 dataset
# Build DESeq2 object and filter low-count genes

```{r create_dds}
# ---- create_dds ----

library(DESeq2)

# Build DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = counts_clean,
  colData   = meta_aligned,
  design    = ~ condition
)

# Remove genes with very low counts
dds <- dds[rowSums(counts(dds)) > 10, ]

dds

```
## 8. Differential expression model
# Run DESeq2 and list available contrasts


```{r run_deseq}
# ---- run_deseq ----

# Run DESeq2
dds <- DESeq(dds)

# Extract results for all conditions vs reference
# By default, reference = first level of condition
resultsNames(dds)

```

```{r NAFL vs control}
# ---- results_NAFL_vs_control ----

res_nafl <- results(dds, contrast = c("condition", "NAFL", "control"))
summary(res_nafl)

```


```{r NASH_F2 vs NASH_F0-F1}
# ---- results_NASH_F2_vs_F0F1 ----

res_f2 <- results(dds, contrast = c("condition", "NASH_F2", "NASH_F0-F1"))
summary(res_f2)

```
## 9. Extract results for main contrast (NASH F2 vs F0–F1)
# Annotate genes and classify DEGs

```{r extract_results_F2_vs_F0F1 }
# ---- extract_results_F2_vs_F0F1 ----

library(tidyverse)

res_df <- as.data.frame(res_f2) %>%
  rownames_to_column("ensembl_id") %>%
  filter(!is.na(padj))

# Add gene annotation
library(org.Hs.eg.db)
library(AnnotationDbi)

gene_anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = res_df$ensembl_id,
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "SYMBOL", "GENENAME", "ENTREZID")
) %>% distinct(ENSEMBL, .keep_all = TRUE)

res_ann <- res_df %>%
  left_join(gene_anno, by = c("ensembl_id" = "ENSEMBL")) %>%
  mutate(
    label = if_else(!is.na(SYMBOL), SYMBOL, ensembl_id),
    direction = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "UP",
      padj < 0.05 & log2FoldChange < -1 ~ "DOWN",
      TRUE ~ "NS"
    ),
    neg_log10_padj = -log10(padj)
  )

head(res_ann)
table(res_ann$direction)

```
## 10. PCA for global sample structure
# Visualize sample clustering using VST-transformed data


```{r vst_pca}
# ---- vst_pca ----

vst_data <- vst(dds, blind = FALSE)

pca_data <- plotPCA(vst_data, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

library(ggplot2)

ggplot(pca_data, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  labs(
    title = "PCA — NASH_F2 vs NASH_F0-F1",
    x = paste0("PC1 (", percentVar[1], "%)"),
    y = paste0("PC2 (", percentVar[2], "%)")
  ) +
  theme_bw()

```
## 11. Sample-to-sample distance heatmap
# Assess global similarity between samples


```{r sample_distances}
# ---- sample_distances ----

library(pheatmap)

dist_mat <- dist(t(assay(vst_data)))
dist_mat <- as.matrix(dist_mat)

pheatmap(
  dist_mat,
  annotation_col = meta_aligned["condition"],
  main = "Sample-to-sample distances"
)

```
## 12. Volcano plot (NASH F2 vs F0–F1)
# Visualize significant DEGs

```{r volcano_F2_vs_F0F1}
# ---- volcano_F2_vs_F0F1 ----

library(ggrepel)

ggplot(res_ann, aes(log2FoldChange, neg_log10_padj, color = direction)) +
  geom_point(alpha = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  scale_color_manual(values = c(UP = "red", DOWN = "blue", NS = "grey")) +
  labs(
    title = "Volcano — NASH_F2 vs NASH_F0-F1",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  theme_bw()

```

## 13. Heatmap of top 50 DEGs
# Highlight most significant genes in the contrast

```{r heatmap_top50_F2_vs_F0F1 }
# ---- heatmap_top50_F2_vs_F0F1 ----

top50 <- res_ann %>%
  filter(direction != "NS") %>%
  arrange(padj) %>%
  slice_head(n = 50) %>%
  pull(ensembl_id)

vst_mat <- assay(vst_data)[top50, ]
vst_mat <- t(scale(t(vst_mat)))

pheatmap(
  vst_mat,
  annotation_col = meta_aligned["condition"],
  show_rownames = TRUE,
  main = "Top 50 DEGs — NASH_F2 vs NASH_F0-F1"
)

```
## 14. GO enrichment analysis (F2 vs F0–F1)
# Identify enriched biological processes


```{r enrichment_F2_vs_F0F1}
# ---- enrichment_F2_vs_F0F1 ----

library(clusterProfiler)
library(enrichplot)

up_genes <- res_ann %>% filter(direction == "UP", !is.na(ENTREZID)) %>% pull(ENTREZID)
down_genes <- res_ann %>% filter(direction == "DOWN", !is.na(ENTREZID)) %>% pull(ENTREZID)
universe <- res_ann %>% filter(!is.na(ENTREZID)) %>% pull(ENTREZID)

go_up <- enrichGO(
  gene = up_genes,
  universe = universe,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  readable = TRUE
)

dotplot(go_up, showCategory = 20) + ggtitle("GO — Upregulated (F2 vs F0-F1)")

```
### Recap — NASH F2 vs F0–F1

- **DEGs:** 27 upregulated, 2 downregulated  
- **PCA:** low variance explained (≈14% + 10%), consistent with subtle biological transition  
- **Volcano:** few significant genes, but clear immune-related hits  
- **Heatmap:** moderate separation, expected due to patient heterogeneity  
- **GO enrichment:** strong immune signature  
  - Response to bacteria (GeneRatio ≈ 0.275)  
  - Antimicrobial humoral response (≈0.225)  
  - Chemotaxis (≈0.225)  

**Interpretation:** Early fibrotic activation driven by immune recruitment and antimicrobial signaling.



## 15. Extract results for secondary contrast (NAFL vs Control)
# Annotate genes and classify DEGs

```{r extract_results_NAFL_vs_control}
# ---- extract_results_NAFL_vs_control ----

library(tidyverse)

res_df_nafl <- as.data.frame(res_nafl) %>%
  rownames_to_column("ensembl_id") %>%
  filter(!is.na(padj))

# Annotation
library(org.Hs.eg.db)
library(AnnotationDbi)

gene_anno <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = res_df_nafl$ensembl_id,
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "SYMBOL", "GENENAME", "ENTREZID")
) %>% distinct(ENSEMBL, .keep_all = TRUE)

res_nafl_ann <- res_df_nafl %>%
  left_join(gene_anno, by = c("ensembl_id" = "ENSEMBL")) %>%
  mutate(
    label = if_else(!is.na(SYMBOL), SYMBOL, ensembl_id),
    direction = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "UP",
      padj < 0.05 & log2FoldChange < -1 ~ "DOWN",
      TRUE ~ "NS"
    ),
    neg_log10_padj = -log10(padj)
  )

table(res_nafl_ann$direction)

```
## 16. Volcano plot (NAFL vs Control)
# Visualize large-scale metabolic dysregulation

```{r volcano_NAFL_vs_control}
# ---- volcano_NAFL_vs_control ----

library(ggrepel)

ggplot(res_nafl_ann, aes(log2FoldChange, neg_log10_padj, color = direction)) +
  geom_point(alpha = 0.6) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  scale_color_manual(values = c(UP = "red", DOWN = "blue", NS = "grey")) +
  labs(
    title = "Volcano — NAFL vs control",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  theme_bw()

```
## 17. Heatmap of top 50 DEGs (NAFL vs Control)
# Show major metabolic gene shifts

```{r heatmap_top50_NAFL_vs_control}
# ---- heatmap_top50_NAFL_vs_control ----

top50_nafl <- res_nafl_ann %>%
  filter(direction != "NS") %>%
  arrange(padj) %>%
  slice_head(n = 50) %>%
  pull(ensembl_id)

vst_mat <- assay(vst_data)[top50_nafl, ]
vst_mat <- t(scale(t(vst_mat)))

pheatmap(
  vst_mat,
  annotation_col = meta_aligned["condition"],
  show_rownames = TRUE,
  main = "Top 50 DEGs — NAFL vs control"
)

```
## 18. GO enrichment analysis (NAFL vs Control)
# Identify metabolic pathways altered in NAFL

```{r enrichment_NAFL_vs_control}
# ---- enrichment_NAFL_vs_control ----

library(clusterProfiler)
library(enrichplot)

up_nafl <- res_nafl_ann %>% filter(direction == "UP", !is.na(ENTREZID)) %>% pull(ENTREZID)
down_nafl <- res_nafl_ann %>% filter(direction == "DOWN", !is.na(ENTREZID)) %>% pull(ENTREZID)
universe_nafl <- res_nafl_ann %>% filter(!is.na(ENTREZID)) %>% pull(ENTREZID)

go_up_nafl <- enrichGO(
  gene = up_nafl,
  universe = universe_nafl,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  readable = TRUE
)

dotplot(go_up_nafl, showCategory = 20) + ggtitle("GO — Upregulated (NAFL vs control)")

```
### Recap — NAFL vs Control

- **DEGs:** 1102 upregulated, 1838 downregulated  
- **Volcano:** dense and highly significant metabolic signal  
- **Heatmap:** strong separation between control and NAFL  
- **GO enrichment:** metabolic reprogramming  
  - Lipid metabolic regulation (≈25/30 genes, GeneRatio ≈ 0.045)  
  - Cholesterol biosynthesis (≈10/15 genes, GeneRatio ≈ 0.015)  
  - Sterol/secondary alcohol biosynthesis (≈10/15 genes, GeneRatio ≈ 0.015)  

**Interpretation:** Massive metabolic shift consistent with early NAFLD.


# Final Summary

## Two-step model of NAFLD progression

1. **Metabolic dysregulation (NAFL stage)**  
   - Large number of DEGs  
   - Strong enrichment for lipid and cholesterol metabolism  
   - Clear separation between healthy and NAFL samples  

2. **Immune activation (transition to fibrotic NASH)**  
   - Few DEGs but highly specific pathways  
   - Antimicrobial response, chemotaxis, humoral immunity  
   - Matches early fibrogenesis described in the literature  

## Visual Recap Table

```{r summary_table}
# ---- summary_table ----
library(knitr)
library(kableExtra)

summary_df <- data.frame(
  Contrast = c("NAFL vs Control", "NASH F2 vs F0-F1"),
  Up_DEGs = c(1102, 27),
  Down_DEGs = c(1838, 2),
  Main_Theme = c("Metabolic reprogramming", "Immune activation"),
  Key_Pathways = c(
    "Lipid metabolism, cholesterol biosynthesis",
    "Antimicrobial response, chemotaxis"
  )
)

kable(summary_df, format = "html", caption = "Summary of Differential Expression Results") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

```{r summary_plot}
# ---- summary_plot ----
library(ggplot2)

summary_plot_df <- data.frame(
  Contrast = c("NAFL vs Control", "NASH F2 vs F0-F1"),
  Up = c(1102, 27),
  Down = c(1838, 2)
)

summary_plot_df_long <- summary_plot_df %>%
  tidyr::pivot_longer(cols = c("Up", "Down"), names_to = "Direction", values_to = "Count")

ggplot(summary_plot_df_long, aes(x = Contrast, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
  labs(
    title = "Differentially Expressed Genes per Contrast",
    y = "Number of DEGs",
    x = ""
  ) +
  theme_minimal(base_size = 14)

```

