---
title: "NAFLD Transcriptomic Analysis"
author: "Nicolas Legrand"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    highlight: tango
    code_folding: hide
    smooth_scroll: true
---

### Bulk RNA‑seq differential expression across fibrosis stages  
*Author: Nicolas Legrand — Date: `r format(Sys.time(), '%d %B %Y')`*

Non‑alcoholic fatty liver disease (NAFLD) represents a progressive spectrum of liver disorders ranging from simple steatosis to non‑alcoholic steatohepatitis (NASH) and advanced fibrosis. Understanding the molecular mechanisms driving this progression is essential for identifying biomarkers and therapeutic targets.

This report presents a complete RNA‑seq analysis workflow applied to the GEO dataset **GSE135251**, which includes liver biopsy samples spanning multiple fibrosis stages. The analysis covers dataset retrieval, preprocessing, differential expression, visualization, and functional enrichment. The goal is to provide a clear, reproducible, and interpretable transcriptomic overview of NAFLD progression.



```{r chunk_setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  results = "hold"
)

```

## 1. Dataset loading and initial exploration

This section retrieves the GEO dataset GSE135251 and performs an initial inspection of the expression matrix and sample metadata. The goal is to verify data integrity, sample count, and available annotations before processing.

```{r setup}
# ---- Setup ----
library(GEOquery)
library(tidyverse)

# ---- Download GEO dataset ----
gse <- getGEO("GSE135251", GSEMatrix = TRUE)

# Verify download
length(gse)
names(gse)

expr <- exprs(gse[[1]])
meta <- pData(gse[[1]])

dim(expr)
head(expr[,1:5])
head(meta)

```


## 2. Download and extract raw count files

Here we download the raw featureCounts files associated with the dataset and extract them locally. These files contain the raw gene‑level read counts required for downstream differential expression analysis.

```{r download raw, echo=FALSE, results="hide"}
library(GEOquery)

# Télécharger les fichiers supplémentaires (RAW)
getGEOSuppFiles("GSE135251")

# Décompresser l’archive RAW dans un sous-dossier
untar("GSE135251/GSE135251_RAW.tar", exdir = "GSE135251/raw")

# Lister les fichiers disponibles
list.files("GSE135251/raw")

```

## 3. Merge raw count files into a single count matrix

This section consolidates all individual featureCounts outputs into a unified count matrix. Each row corresponds to a gene and each column to a sample. This step ensures consistent gene identifiers and prepares the data for cleaning.

```{r merge_counts}
# ---- merge_counts_fixed ----

library(tidyverse)

files <- list.files("GSE135251/raw", full.names = TRUE)

read_count_file <- function(f) {
  df <- read.delim(f, header = TRUE)
  
  gene_col <- grep("Geneid|gene|Gene|ID", colnames(df), ignore.case = TRUE)
  if (length(gene_col) == 0) gene_col <- 1
  
  count_col <- grep("count|Counts|read", colnames(df), ignore.case = TRUE)
  if (length(count_col) == 0) count_col <- 2
  
  df <- df[, c(gene_col[1], count_col[1])]
  colnames(df) <- c("Geneid", basename(f))
  
  return(df)
}

count_list <- lapply(files, read_count_file)

counts_mat <- Reduce(function(x, y) merge(x, y, by = "Geneid", all = TRUE), count_list)

counts_mat[is.na(counts_mat)] <- 0

rownames(counts_mat) <- counts_mat$Geneid
counts_mat <- counts_mat[, -1]

dim(counts_mat)
counts_mat[1:5, 1:5]

```

## 4. Clean count matrix

We remove technical rows generated by featureCounts (e.g., summary statistics) and convert the matrix into a clean integer format suitable for DESeq2. This ensures that only biologically meaningful gene counts are retained.

```{r clean_counts}
# ---- clean_counts ----

bad_rows <- grep("^__", rownames(counts_mat))
length(bad_rows)

counts_clean <- counts_mat[-bad_rows, ]

counts_clean <- as.matrix(counts_clean)
mode(counts_clean) <- "integer"

dim(counts_clean)
counts_clean[1:5, 1:5]


```

## 5. Clean and format metadata

This section extracts relevant clinical annotations (group, fibrosis stage, NAS score) and formats them into a structured metadata table. A clean condition factor is created to define the experimental groups used in the DESeq2 model.

```{r clean_metadata}

# ---- clean_metadata ----

library(dplyr)
library(stringr)

# Start from meta
meta_clean <- meta %>%
  dplyr::select(`group in paper:ch1`, `fibrosis stage:ch1`, `nas score:ch1`)

# Rename columns for clarity
meta_clean <- meta_clean %>%
  dplyr::rename(
    group = `group in paper:ch1`,
    fibrosis_stage = `fibrosis stage:ch1`,
    nas_score = `nas score:ch1`
  )

# Extract condition from group
meta_clean$condition <- meta_clean$group

# Make condition a factor
meta_clean$condition <- factor(meta_clean$condition)

# Check
table(meta_clean$condition)
head(meta_clean)


```

```{r clean_colnames}
# ---- clean_colnames ----

# Extract GSM ID from filenames
clean_names <- colnames(counts_clean) %>%
  stringr::str_extract("GSM[0-9]+")

# Replace column names
colnames(counts_clean) <- clean_names

# Check
head(colnames(counts_clean))

```
## 6. Align metadata with count matrix

Here we ensure that sample names match perfectly between the count matrix and metadata. Proper alignment is essential for building a valid DESeq2 dataset and avoiding downstream inconsistencies.

```{r check_alignment}
# ---- check_alignment2 ----

# Check again
all(colnames(counts_clean) == rownames(meta_clean))

# Align metadata to counts
meta_aligned <- meta_clean[colnames(counts_clean), ]

# Check
all(colnames(counts_clean) == rownames(meta_aligned))

head(meta_aligned)

```

## 7. Create DESeq2 dataset

We construct the DESeq2 object using the cleaned count matrix and aligned metadata. Low‑count genes are filtered out to improve statistical power and reduce noise.

```{r create_dds}
# ---- create_dds ----

library(DESeq2)

# Build DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = counts_clean,
  colData   = meta_aligned,
  design    = ~ condition
)

# Remove genes with very low counts
dds <- dds[rowSums(counts(dds)) > 10, ]

dds

```
## 8. Differential expression model

This section runs the DESeq2 pipeline to estimate size factors, dispersions, and model coefficients. It also lists the available contrasts corresponding to the different fibrosis stages.

```{r run_deseq}
# ---- run_deseq ----

# Run DESeq2
dds <- DESeq(dds)

# Extract results for all conditions vs reference
# By default, reference = first level of condition
resultsNames(dds)

```
```{r vst_transform}
# ---- vst_transform ----

vst_data <- vst(dds, blind = FALSE)
```

## 9. Multi‑contrast differential expression analysis

This section performs a complete multi‑contrast analysis across all fibrosis stages. For each contrast, we generate:

annotated differential expression results

volcano plots

PCA projections

heatmaps of strongly regulated genes

GO enrichment of upregulated genes

The goal is to provide a comprehensive transcriptional overview of NAFLD progression.

```{r all_contrasts}
# ---- all_contrasts ----

contrast_list <- list(
  NAFL_vs_control   = c("condition", "NAFL",       "control"),
  F0F1_vs_control   = c("condition", "NASH_F0-F1", "control"),
  F2_vs_control     = c("condition", "NASH_F2",    "control"),
  F3_vs_control     = c("condition", "NASH_F3",    "control"),
  F4_vs_control     = c("condition", "NASH_F4",    "control")
)

all_results <- list()

for (name in names(contrast_list)) {
  cat("Running:", name, "\n")
  res <- results(dds, contrast = contrast_list[[name]])
  res <- as.data.frame(res)
  res$ensembl_id <- rownames(res)
  all_results[[name]] <- res
}
```

### 9.2 Annotation function

This function enriches DESeq2 results with gene symbols, descriptions, and ENTREZ IDs, enabling clearer interpretation and downstream functional analysis.

```{r annotate_function}
# ---- annotate_function ----

library(dplyr)
library(org.Hs.eg.db)
library(AnnotationDbi)

annotate_results <- function(res_df) {

  gene_anno <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = res_df$ensembl_id,
    keytype = "ENSEMBL",
    columns = c("ENSEMBL", "SYMBOL", "GENENAME", "ENTREZID")
  ) |> distinct(ENSEMBL, .keep_all = TRUE)

  res_ann <- res_df |>
    left_join(gene_anno, by = c("ensembl_id" = "ENSEMBL")) |>
    mutate(
      label = if_else(!is.na(SYMBOL), SYMBOL, ensembl_id),
      direction = case_when(
        padj < 0.05 & log2FoldChange >  1 ~ "UP",
        padj < 0.05 & log2FoldChange < -1 ~ "DOWN",
        TRUE ~ "NS"
      ),
      neg_log10_padj = -log10(padj)
    )

  return(res_ann)
}

```

### 9.3 Volcano plot function

This function generates volcano plots highlighting significantly up‑ and down‑regulated genes, providing an intuitive overview of differential expression patterns.

```{r volcano_function}
# ---- volcano_function ----

library(ggplot2)
library(ggrepel)

plot_volcano <- function(res_ann, title) {
  ggplot(res_ann, aes(log2FoldChange, neg_log10_padj, color = direction)) +
    geom_point(alpha = 0.6) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    scale_color_manual(values = c(UP="red", DOWN="blue", NS="grey")) +
    labs(title = paste("Volcano —", title),
         x = "log2 Fold Change",
         y = "-log10 adjusted p-value") +
    theme_bw()
}

```

### 9.4 PCA function (per contrast)

This PCA visualizes sample clustering for each contrast, helping assess group separation and potential batch effects.

```{r pca_function}
# ---- pca_function ----

library(ggplot2)

plot_pca_contrast <- function(vst_data, meta_aligned, group1, group2, title) {

  # 1. Sélectionner uniquement les échantillons du contraste
  samples_keep <- rownames(meta_aligned)[meta_aligned$condition %in% c(group1, group2)]
  vst_sub <- vst_data[, samples_keep]

  # 2. Extraire matrice VST
  mat <- t(assay(vst_sub))

  # 3. Retirer les colonnes constantes ou quasi-constantes
  vars <- apply(mat, 2, var)
  mat <- mat[, vars > 1e-6, drop = FALSE]

  # 4. Vérifier qu'il reste assez de colonnes pour une PCA
  if (ncol(mat) < 2) {
    return(list(plot = NULL, pca = NULL, percentVar = NULL))
  }

  # 5. PCA
  pca <- prcomp(mat, scale. = TRUE)
  percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

  # 6. Dataframe pour ggplot
  df <- data.frame(
    PC1 = pca$x[,1],
    PC2 = pca$x[,2],
    condition = meta_aligned[samples_keep, "condition"]
  )

  # 7. Plot PCA
  p <- ggplot(df, aes(PC1, PC2, color = condition)) +
    geom_point(size = 3) +
    labs(
      title = paste("PCA —", title),
      x = paste0("PC1 (", percentVar[1], "%)"),
      y = paste0("PC2 (", percentVar[2], "%)")
    ) +
    theme_bw()

  return(list(plot = p, pca = pca, percentVar = percentVar))
}

```

### 9.5 Heatmap function (strong DEGs)

This function displays the top polarized genes for each contrast, offering a focused view of the most biologically relevant transcriptional changes.

```{r heatmap_function}
# ---- heatmap_function ----
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tibble)

plot_heatmap <- function(res_ann, vst_data, meta_aligned, title) {

  # Gènes fortement modulés (polarisation stricte)
  strong_genes <- res_ann |>
    filter(direction != "NS") |>
    filter(padj < 0.01, abs(log2FoldChange) >= 2) |>
    arrange(padj) |>
    slice_head(n = 20) |>
    pull(ensembl_id)

  if (length(strong_genes) < 5) {
    cat("Not enough strong DEGs for heatmap:", title, "\n")
    return(NULL)
  }

  vst_mat <- assay(vst_data)[strong_genes, ]
  vst_mat <- t(scale(t(vst_mat)))

  row_anno <- res_ann |>
    filter(ensembl_id %in% strong_genes) |>
    select(ensembl_id, direction) |>
    distinct() |>
    column_to_rownames("ensembl_id")

  my_palette <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(255)

  pheatmap(
    vst_mat,
    annotation_col = meta_aligned["condition"],
    annotation_row = row_anno,
    show_rownames = TRUE,
    show_colnames = FALSE,
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    color = my_palette,
    fontsize_row = 8,
    main = paste("Strong polarized DEGs —", title)
  )
}


```

### 9.6 GO enrichment function

This step identifies enriched biological processes among upregulated genes, providing functional interpretation of the transcriptional signatures.

```{r go_function}
# ---- go_function ----

library(clusterProfiler)
library(enrichplot)

run_go <- function(res_ann, title) {

  up_ids <- res_ann |>
    filter(direction == "UP", !is.na(ENTREZID)) |>
    pull(ENTREZID)

  universe_ids <- res_ann |>
    filter(!is.na(ENTREZID)) |>
    pull(ENTREZID)

  if (length(up_ids) < 5) {
    cat("Not enough genes for GO:", title, "\n")
    return(NULL)
  }

  go_res <- enrichGO(
    gene          = up_ids,
    universe      = universe_ids,
    OrgDb         = org.Hs.eg.db,
    ont           = "BP",
    pAdjustMethod = "BH",
    readable      = TRUE
  )

  dotplot(go_res, showCategory = 15) +
    ggtitle(paste("GO — Upregulated genes:", title))
}

```

### 9.7 Run full multi‑contrast pipeline

This step identifies enriched biological processes among upregulated genes, providing functional interpretation of the transcriptional signatures.

```{r run_pipeline}
# ---- run_pipeline ----

annotated_results <- list()

for (name in names(all_results)) {

  cat("\n\n===== Processing:", name, "=====\n\n")

  # 1. Annoter
  res_ann <- annotate_results(all_results[[name]])
  annotated_results[[name]] <- res_ann


  # ============================================================
  # 2. VOLCANO + INTERPRÉTATION
  # ============================================================

  print(plot_volcano(res_ann, name))

  up_count   <- sum(res_ann$direction == "UP",   na.rm = TRUE)
  down_count <- sum(res_ann$direction == "DOWN", na.rm = TRUE)

  cat("\n**Volcano interpretation —", name, "**\n")
  cat("- Upregulated genes:", up_count, "\n")
  cat("- Downregulated genes:", down_count, "\n")

  if ((up_count + down_count) < 50) {
    cat("→ Weak differential expression signal.\n")
  } else if (up_count > down_count) {
    cat("→ Upregulation dominates.\n")
  } else {
    cat("→ Downregulation dominates.\n")
  }



  # ============================================================
  # 3. PCA + INTERPRÉTATION
  # ============================================================

  # ---- PCA block inside run_pipeline ----

  groups <- contrast_list[[name]][2:3]
  pca_out <- plot_pca_contrast(vst_data, meta_aligned, groups[1], groups[2], name)
  
  # Affichage du plot PCA
  print(pca_out$plot)
  
  # Interprétation PCA
  cat("\n**PCA interpretation —", name, "**\n")
  
  if (is.null(pca_out$pca)) {
  
    cat("→ PCA could not be computed (insufficient variance).\n")
  
  } else {
  
    pc1 <- pca_out$percentVar[1]
    pc2 <- pca_out$percentVar[2]
  
    cat("- PC1 explains:", pc1, "% of variance\n")
    cat("- PC2 explains:", pc2, "% of variance\n")
  
    if (pc1 < 10) {
      cat("→ Very weak structure: groups likely overlap.\n")
    } else if (pc1 < 20) {
      cat("→ Moderate structure: partial separation.\n")
    } else {
      cat("→ Strong structure: clear separation likely.\n")
    }
  
    cat("→ Visual separation on the PCA plot indicates biological divergence.\n")
  }




  # ============================================================
  # 4. HEATMAP + INTERPRÉTATION
  # ============================================================

  plot_heatmap(res_ann, vst_data, meta_aligned, name)

  strong_genes_n <- sum(res_ann$padj < 0.01 & abs(res_ann$log2FoldChange) >= 2, na.rm = TRUE)

  cat("\n**Heatmap interpretation —", name, "**\n")

  if (strong_genes_n < 10) {
    cat("→ Few strongly polarized genes: weak transcriptional signature.\n")
  } else if (strong_genes_n < 50) {
    cat("→ Moderate number of polarized genes: interpretable patterns.\n")
  } else {
    cat("→ Strong polarized signature: clear up/down blocks visible.\n")
  }

  cat("→ Red = upregulated, Blue = downregulated.\n")
  cat("→ Clear block structure indicates coherent biological programs.\n")



  # ============================================================
  # 5. GO + INTERPRÉTATION
  # ============================================================

  go_plot <- run_go(res_ann, name)
  print(go_plot)

  cat("\n**GO interpretation —", name, "**\n")
  cat("→ Enriched GO terms highlight the dominant biological processes.\n")
  cat("→ Typically immune activation, ECM remodeling, or metabolic pathways in NAFLD.\n")
}


```

### 9.8 Summary table of all contrasts

A concise table summarizing the number of up‑ and down‑regulated genes for each contrast.

```{r summary_all}
# ---- summary_all ----

library(knitr)
library(kableExtra)

summary_df <- data.frame(
  Contrast  = names(annotated_results),
  Up_DEGs   = sapply(annotated_results, function(x) sum(x$direction == "UP")),
  Down_DEGs = sapply(annotated_results, function(x) sum(x$direction == "DOWN"))
)

kable(summary_df, format="html", caption="Summary of all contrasts") %>%
  kable_styling(full_width=FALSE, bootstrap_options=c("striped","hover"))

```

### 9.9 Summary barplot of DEGs per contrast

A stacked barplot visualizing the distribution of DEGs across contrasts.

```{r summary_barplot}
# ---- summary_barplot ----

library(ggplot2)
library(tidyr)

summary_long <- summary_df |>
  pivot_longer(cols = c("Up_DEGs", "Down_DEGs"),
               names_to = "Direction", values_to = "Count")

ggplot(summary_long, aes(x = Contrast, y = Count, fill = Direction)) +
  geom_bar(stat="identity", position="stack") +
  scale_fill_manual(values=c(Up_DEGs="red", Down_DEGs="blue")) +
  theme_minimal(base_size=14) +
  labs(title="DEGs per contrast", y="Number of DEGs", x="")

```
